# Exploiting the cell ontology

```{r, echo=FALSE, results='asis'}
library(rebook)
chapterPreamble(TRUE)
```

## Motivation

As previously discussed in Section \@ref(using-harmonized-labels),
`r Biocpkg("SingleR")` maps the labels in its references to the [Cell Ontology](https://www.ebi.ac.uk/ols/ontologies/cl).
The most obvious advantage of doing this is to provide a standardized vocabulary with which to describe cell types,
thus facilitating integrated analyses with multiple references.
However, another useful feature of the Cell Ontology is its hierarchical organization of terms,
allowing us to adjust cell type annotations to the desired resolution.
This represents a more dynamic alternative to the static `label.main` and `label.fine` options in each reference.

## Basic manipulation

We use the `r Biocpkg("ontoProc")` package to load in the Cell Ontology.
This produces an `ontology_index` object (from the `r CRANpkg("ontologyIndex")` package)
that we can query for various pieces of information.

```{r}
# TODO: wrap in utility function.
library(ontoProc)
bfc <- BiocFileCache::BiocFileCache(ask=FALSE)
path <- BiocFileCache::bfcrpath(bfc, "http://purl.obolibrary.org/obo/cl.obo")
cl <- get_ontology(path, extract_tags="everything")
cl
```

The most immediate use of this object lies in mapping ontology terms to their plain-English descriptions.
We can use this to translate annotations produced by `SingleR()` from the `label.ont` labels into a more interpretable form.
We demonstrate this approach using `r Biocpkg("celldex")`'s collection of mouse RNA-seq references [@aran2019reference].

```{r}
head(cl$name) # short name
head(cl$def) # longer definition

library(celldex)
ref <- MouseRNAseqData(cell.ont="nonna")
translated <- cl$name[ref$label.ont]
head(translated)
```

Another interesting application involves examining the relationship between different terms.
The ontology itself is a directed acyclic graph, so we can can convert it into `graph` object
for advanced queries using the `r CRANpkg("igraph")` package.
Each edge represents an "is a" relationship where each vertex represents a specialized case of the concept of the parent node.

```{r}
# TODO: wrap in utility function.
parents <- cl$parents
self <- rep(names(parents), lengths(parents))

library(igraph)
g <- make_graph(rbind(unlist(parents), self))
g
```

One query involves identifying all descendents of a particular term of interest.
This can be useful when searching for a cell type in the presence of variable annotation resolution;
for example, a search for "epithelial cell" can be configured to pick up all child terms 
such as "endothelial cell" and "ependymal cell".

```{r}
term <- "CL:0000624"
cl$name[term]
all.kids <- names(subcomponent(g, term))
head(cl$name[all.kids])
```

Alternatively, we might be interested in the last common ancestor (LCA) for a set of terms.
This is the furthest term - or, in some cases, multiple terms - from the root of the ontology
that is also an ancestor of all of the terms of interest.
We will use this LCA concept in the next section to adjust resolution across multiple references. 

```{r}
terms <- c("CL:0000624", "CL:0000785", "CL:0000623")
cl$name[terms]

# TODO: god, put this in a function somewhere.
all.ancestors <- lapply(terms, subcomponent, graph=g, mode="in")
all.ancestors <- lapply(all.ancestors, names)
common.ancestors <- Reduce(intersect, all.ancestors)

ancestors.of.ancestors <- lapply(common.ancestors, subcomponent, graph=g, mode="in")
ancestors.of.ancestors <- lapply(ancestors.of.ancestors, names)
ancestors.of.ancestors <- mapply(setdiff, ancestors.of.ancestors, common.ancestors) 

latest.common.ancestors <- setdiff(common.ancestors, unlist(ancestors.of.ancestors))
cl$name[latest.common.ancestors]
```

## Adjusting resolution

We can use the ontology graph to adjust the resolution of the reference labels by rolling up overly-specific terms to their LCA.
The `findCommonAncestors()` utility takes a set of terms and returns a list of potential LCAs for various subsets of those terms.
Users can inspect this list to identify LCAs at the desired resolution and then map their descendent terms to those LCAs.

```{r}
findCommonAncestors <- function(..., g, remove.self=TRUE, names=NULL) {
    terms <- list(...)
    if (is.null(names(terms))) {
        names(terms) <- sprintf("set%i", seq_along(terms))
    }

    all.terms <- unique(unlist(terms))
    all.ancestors <- lapply(all.terms, subcomponent, graph=g, mode="in")
    all.ancestors <- lapply(all.ancestors, names)
    by.ancestor <- split(
        rep(all.terms, lengths(all.ancestors)),
        unlist(all.ancestors)
    )

    # Removing ancestor nodes with the same count as its children.
    available <- names(by.ancestor)
    for (i in available) {
        if (!i %in% names(by.ancestor)) {
            next
        }

        counts <- lengths(by.ancestor)
        cur.ancestors <- subcomponent(g, i, mode="in")
        cur.ancestors <- setdiff(names(cur.ancestors), i)
        drop <- cur.ancestors[counts[i]==counts[cur.ancestors]]
        by.ancestor <- by.ancestor[!names(by.ancestor) %in% drop]
    }

    if (remove.self) {
        by.ancestor <- by.ancestor[lengths(by.ancestor) > 1L]
    }
    by.ancestor <- by.ancestor[order(lengths(by.ancestor))] # most specific terms first.

    # Decorating the output.
    for (i in names(by.ancestor)) {
        current <- by.ancestor[[i]]
        df <- DataFrame(row.names=current)

        curout <- list()
        if (!is.null(names)) {
            curout$name <- unname(names[i])
            df$name <- names[current]
        }

        presence <- list()
        for (b in names(terms)) {
            presence[[b]] <- current %in% terms[[b]]
        }
        df <- cbind(df, do.call(DataFrame, presence))

        curout$descendents <- df
        by.ancestor[[i]] <- curout
    }

    by.ancestor
}

lca <- findCommonAncestors(ref$label.ont, g=g, names=cl$name)
head(lca)
```

We can also use this function to synchronize multiple sets of terms to the same resolution.
Here, we consider the ImmGen dataset [@ImmGenRef], which provides highly resolved annotation of immune cell types.
The `findCommonAncestors()` function specifies the origins of the descendents for each LCA,
allowing us to focus on LCAs that have representatives in both sets of terms.

```{r}
ref2 <- ImmGenData(cell.ont="nonna")
lca2 <- findCommonAncestors(MouseRNA=ref$label.ont,
    ImmGen=ref2$label.ont, g=g, names=cl$name)
head(lca2)
```

For example, we might notice that the mouse RNA-seq reference only has a single "T cell" term.
To synchronize resolution across references, 
we would need to roll up all of the ImmGen's finely resolved subsets into that LCA as shown below.
Of course, this results in some loss of precision and information;
whether this is an acceptable price for simpler interpretation is a decision that is left to the user.

```{r}
children <- lca2$`CL:0000084`$descendents
children

# Synchronization:
synced.mm <- ref$label.ont
synced.mm[synced.mm %in% rownames(children)] <- "CL:0000084"
synced.ig <- ref2$label.ont
synced.ig[synced.ig %in% rownames(children)] <- "CL:0000084"
```

## Session information {-}

```{r, echo=FALSE, results="asis"}
prettySessionInfo()
```
