---
bibliography: ref.bib
---

# (PART) Workflows {-}

# Cross-annotating pancreas

```{r, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble(cache=TRUE)
```

## Loading the data

We load the @muraro2016singlecell dataset as our reference, 
removing unlabelled cells or cells without a clear label.
We also need to compute log-expression values for use in `SingleR()`.

```{r loading-muraro}
library(scRNAseq)
sceM <- MuraroPancreasData()

sceM <- sceM[,!is.na(sceM$label) & sceM$label!="unclear"] 

library(scater)
sceM <- logNormCounts(sceM)
```

```{r}
# Examining the distribution of labels in this reference.
table(sceM$label)
```

We load the @grun2016denovo dataset as our test,
applying some basic quality control to remove low-quality cells in some of the batches
(see [here](https://osca.bioconductor.org/grun-human-pancreas-cel-seq2.html#quality-control-8) for details).
Technically speaking, this does not need log-expression values but we compute them anyway for convenience.

```{r loading-grun}
sceG <- GrunPancreasData()

sceG <- addPerCellQC(sceG)
qc <- quickPerCellQC(colData(sceG), 
    percent_subsets="altexps_ERCC_percent",
    batch=sceG$donor,
    subset=sceG$donor %in% c("D17", "D7", "D2"))
sceG <- sceG[,!qc$discard]

sceG <- logNormCounts(sceG)
```

```{r}
ncol(sceG)
```

## Applying the annotation

We apply `SingleR()` with Wilcoxon rank sum test-based marker detection to annotate the Grun dataset.

```{r annotation}
library(SingleR)
pred.grun <- SingleR(test=sceG, ref=sceM, labels=sceM$label, de.method="wilcox")
```

We examine the distribution of predicted labels:

```{r}
table(pred.grun$labels)
```

## Diagnostics

## Session information {-}

```{r, results='asis'}
prettySessionInfo()
```
