# Advanced options

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
pred.hesc <- readRDS("pred.hesc.rds")
```

## Improving efficiency

Advanced users can split the `SingleR()` workflow into two separate training and classification steps.
This means that training (e.g., marker detection, assembling of nearest-neighbor indices) only needs to be performed once.
The resulting data structures can then be re-used across multiple classifications with different test datasets, provided the test feature set is identical to or a superset of the features in the training set.
For example:

```{r}
library(scRNAseq)
hESCs <- LaMannoBrainData('human-es')

library(scater)
hESCs <- logNormCounts(hESCs)

library(SingleR)
hpca.se <- HumanPrimaryCellAtlasData()
common <- intersect(rownames(hESCs), rownames(hpca.se))
trained <- trainSingleR(hpca.se[common,], labels=hpca.se$label.main)
pred.hesc2 <- classifySingleR(hESCs[common,], trained)
table(pred.hesc$labels, pred.hesc2$labels)
```

Other efficiency improvements are possible through several arguments:

- Switching to an approximate algorithm for the nearest neighbor search in `trainSingleR()` via the `BNPARAM=` argument from the `r Biocpkg("BiocNeighbors")` package.
- Parallelizing the fine-tuning step in `classifySingleR()` with the `BPPARAM=` argument from the `r Biocpkg("BiocParallel")` package.

These arguments can also be specified in the `SingleR()` command.
